---
title: "MFE Programming Workshop"
author: "Brett Dunn"
date: "Fall 2016"
output:
  beamer_presentation:
    includes:
      in_header: slide_header.tex
    keep_tex: false
subtitle: Week 1
slide_level: 2
---

# Goals
- Learn to program in R.

- What does programming mean?
    - Language syntax
    - Debugging
    - Finding solutions
    - Translating math to code
  
- This is just the beginning, you'll develop these skills throughout the program.

# R vs C++
 
- Both are useful and you will use both in the MFE program
- R is a high-level language
    - Low programmer time
    - It is a great tool for data munging, statistics, regressions, ect.
    - However, certain tasks in R can be slow (e.g. loops) 
- C++ is very fast, but it takes longer to write programs
- We can use both together!
- Write most of the program in R, and convert the slow parts of the program to C++

# Structure
- I will talk at the beginning of each class
- For the remainder of the time you will break into your study groups and work on
  programming tasks 
- Tasks are designed to introduce you to the building blocks that will
  be used for course assignments throughout the MFE program
- This course is a programming course with emphasis on methods for
  finance:
    - You will see finance terms and math
    - You \emph{may} not understand all of the finance, but you will
    learn it throughout the program
- The key skills will be translating mathematical algorithms into code
  and developing the ability to find helpful resources

# Questions
\begin{center}
\Large Any questions before we start?
\end{center}

# R Resources: Books
- Introductory:
    - R for Everyone by Jared P. Lander
    - R Cookbook by Paul Teetor (free at [UCLA LearnIT](https://learnit.ucla.edu/))
- Intermediate:
    - The Art of R Programming by Norman Matloff
- Advanced:
    - Software for Data Analysis by John Chambers
    - Advanced R

# Other Resources
- Book series:
    - [Use R!](http://link.springer.com/bookseries/6991) Springer series
        - FYI: Many Springer textbooks are just $25 through <http://link.springer.com/>.  You need to be on campus or signed into the UCLA VPN. You can download the pdfs for free.
    - O'Reilly R Books (free at [UCLA LearnIT](https://learnit.ucla.edu/))
- Built in documentation!
    - `?funcname`
- Journal of Statistical Software
- Data science courses on [Coursera](https://www.coursera.org/specializations/jhu-data-science)
- Data Camp
- Google, Stack Overflow, ect.

# R Environment
- First, you need an R distribution
    - I recommend Microsoft R Open
    - <https://mran.revolutionanalytics.com/download/>
- Second, you need an integrated development environment (IDE) for R
    - [R Studio](https://www.rstudio.com/products/rstudio/download/preview/) is a fantastic environment to interact with R
    - Other options:
        - [R Tools for Visual Studio](https://www.visualstudio.com/en-us/features/rtvs-vs.aspx) if you use Visual Studio
        - [Emacs Speaks Statistics (ESS)](http://ess.r-project.org/) if you use Emacs
- I am going to assume that you have a working installation of R
  Studio and that you have a basic understanding of how it works
- I will show you some Visual Studio
- My focus is going to be on R programming

# RStudio

\centerline{
  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{./graphics/RStudio.png}
}

# R Tools for Visual Studio

\centerline{
  \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{./graphics/RTVS.png}
}


# Command Line Interface
- To run a command in `R`, type it into the consule next to the `>` symbol and press the Enter key
```{r}
2 + 3
```
- Up Arrow key + Enter repeats the line of code
- Esc (Windows/Mac) or Ctrl-C (Linux) interups a command

# RStudio
- To start, create a new R Script file.
    - File/New File/R Script
- You can type your commands in the R Script file and run them on the Console
    - Easy way to save your work
    -`Ctrl+Enter` sends the line at the cusor to the consul
    -`Ctrl+Shift+S` runs the entite file
    - Help/Keyboard Shortcuts lists all the available shortcuts
        - Check out the multiple cursors
- For larger tasks with many files, create an R project
- Visual Studio is similar

# R Packages
- R has a huge collection of user-contributed packages
    - See CRAN
- A package is essentially a library of prewritten code designed to accomplish some task or a collection of tasks
- Not all packages are of the same quality

# R Packages
- Installing a packages:
    - `Ctrl+7` in RStudio accesses the packages pane
    - You can also type `install.packages("packageName")`
- Uninstalling a package:
    - `remove.packages`
- Loading packages:
    - `require(packageName)` or `library(packageName)` loads a package into R
    - The difference is that `require ` returns `TRUE` if the package loads or `FALSE` if it doesn't.
- Unloading packages
    - `detach("package:packageName")`
- If two packages have the same function name use two colons:
    -`package1::func` or `package2::func`

# R as a language
- R is object oriented
    - Everything is an object and functions operate differently when passed different types of objects
- R is functional
    - You write fewer loops
    - You write cleaner code

# General Comments
- Make your code easy to read
- Check out the google style guide
- Comment your code!

# Variables
- Unlike C++, R does not require variable types to be decleared
- A variable can take on any data type
- A variable can also hold any R object such as a function, the result of an analysis, a plot, ect.
- Variable assignment is done with  `<-`
    - `=` works, but there are reasons to prefer `<-`
- We can remove variables (e.g. to free up memory) with the `rm` function
    ```{r}
    x <- 2
    x
    rm(x)  # removes x
    ```

# Data Types
- The four main types of data most likely to be used are:
    1. `numeric`
    2. `character` (string)
    3. `Date/POSIXct` (time-based)
    4. `logical (TRUE/FALSE)`

- The data type can be checked with the `class` function
```{r}
x <- as.Date("2010-12-21")
class(x)
```

